---
title: "Prioritizing Potential Aquaculture"
format: pdf
author: Lucian Scher
date: 11/26/25
subtitle: EDS 223 Homework 4
editor_options: 
  chunk_output_type: inline
---

# Load packages

```{r, message=FALSE}
library(terra, quiet = TRUE) # Raster data
library(tmap) # Creating Maps
library(sf, quiet = TRUE)
library(testthat, quiet = TRUE) # For testing code chunks
library(rnaturalearth) # For adding land to our eez map

```

# Prepare data

First we load all necessary data and make sure it has the coordinate reference system.

```{r}
# Bathymetry Data
depth <- rast("data/depth.tif")

# Exclusive Economic Zones (eez)
eez <- st_read("data/wc_regions_clean.shp", quiet = TRUE)

# Sea Surface Temperature (sst) 2008-2012

sst <- c(rast("data/average_annual_sst_2012.tif"),
         rast("data/average_annual_sst_2011.tif"),
         rast("data/average_annual_sst_2010.tif"),
         rast("data/average_annual_sst_2009.tif"),
         rast("data/average_annual_sst_2008.tif"))

# Check if CRS match before moving on 

if (crs(eez) == crs(sst) && crs(sst) == crs(depth)) {
  message("The CRS of this data match!")
} else {
  warning("CRS do NOT match. Transforming to match the CRS of SST")
    # Transform eez CRS
  if (crs(eez) != crs(sst)) {
    eez <- st_transform(eez, crs(sst))
  }
  # Transform depth CRS
  if (crs(depth) != crs(sst)) {
    depth <- project(depth, sst)
  }
  message("All CRS now match the SST CRS.")
}

```

# Process data

Next, we need to process the SST and depth data so that they can be combined. In this case the SST and depth data have slightly different resolutions, extents, and positions.

```{r}
# Calculate mean of sst for each year and convert to Celsius
mean_sst <- mean(sst) - 273.15

# Crop to extent of sst and resample to match sst
depth_resampled <- depth |> 
  crop(mean_sst) |> 
  resample(mean_sst, method = "near")

# Test that depth and sst match in crs, resolution and extent
test_that("Depth and SST rasters are aligned", {
  expect_no_error({
    x <- c(depth_resampled, mean_sst)
  })})
```

# Prioritizing Potential Aquaculture Function

The purpose of this function is to take parameters of certain potential aquaculture marine species and visualize the best West Coast Exclusive Economic Zone for their farming.

```{r}
#' @title Mapping Potential Aquaculture for species in West Coast Exclusive Economic Zones
#' @author Lucian Scher
#' @param species_name Name of species (Scientific or Common)
#' @param sst_min Minimum of temperature range (Degrees Celsius)
#' @param sst_max Maximum of temperature range (Degrees Celsius)
#' @param depth_min Minimum of depth range negative for underwater (Meters)
#' @param depth_max Maximum of depth range 0 for sea-level (Meters)
#' *Optional Parameters:* to update SST, depth or EEZ data in case of changes: Default is data used this repository. Will need to re-prepare and re-process if neccessary to change these.
#' @param sst_data Sea Surface Temperature Data
#' @param depth_data Bathymetry data
#' @param eez_data Exclusive Economic Zones
#'
#' @returns Map of EEZ regions colored by amount of suitable area for given species (species name included in map title)

map_species_suitability <- function(species_name, 
                                     sst_min, sst_max, 
                                     depth_min, depth_max,
                                     sst_data = mean_sst,
                                     depth_data = depth_resampled,
                                     eez_data = eez) {
  
# --- Finding Suitable Locations --- #

  # Create reclassification matrices
  sst_rcl <- matrix(c(
    -Inf, sst_min, 0,
    sst_min, sst_max, 1,
    sst_max, Inf, 0
  ), ncol = 3, byrow = TRUE)
  
  depth_rcl <- matrix(c(
    -Inf, depth_min, 0,
    depth_min, depth_max, 1,
    depth_max, Inf, 0
  ), ncol = 3, byrow = TRUE)
  
  # Reclassify to binary suitability
  sst_bin <- classify(sst_data, rcl = sst_rcl)
  depth_bin <- classify(depth_data, rcl = depth_rcl)
  
  # Replicate depth layer to match sst layers
  depth_stack <- rep(depth_bin, nlyr(sst_bin))
  
  # Identify suitable cells
  suitable_cells <- lapp(c(sst_bin, depth_stack), fun = function(sst, depth) {
    sst * depth
  })
  
# --- Determine the most suitable EEZ ---  #

  # Match CRS
  eez_data <- st_transform(eez_data, crs(suitable_cells))
  
  # Mask to EEZ and rasterize
  suitable_masked <- mask(suitable_cells, vect(eez_data))
  eez_rast <- rasterize(vect(eez_data), suitable_masked[[1]], field = "rgn")
  
  # Calculate suitable area per EEZ
  cell_area <- cellSize(suitable_masked, unit = "km")
  suitable_area_rast <- suitable_masked * cell_area
  total_area_per_eez <- zonal(suitable_area_rast, eez_rast, fun = "sum", na.rm = TRUE)
  
  colnames(total_area_per_eez) <- c("Region", "Suitable Area (km2)")
  ranked_eez <- total_area_per_eez[order(-total_area_per_eez$"Suitable Area (km2)"), ]
  ranked_eez$"Suitable Area (km2)" <- round(ranked_eez$"Suitable Area (km2)", 1)
  
  # --- Create map of suitable EEZ regions --- #
  
  # Add suitable area to EEZ polygons
  eez_data$"Suitable Area (km2)" <- setNames(ranked_eez$"Suitable Area (km2)", 
                                              ranked_eez$Region)[eez_data$rgn]
  
  # Get land polygons cropped to EEZ extent with buffer
  land <- ne_countries(scale = "medium", returnclass = "sf")
  eez_bbox <- st_bbox(eez_data)
  eez_bbox[c("xmin", "ymin")] <- eez_bbox[c("xmin", "ymin")] - 2
  eez_bbox[c("xmax", "ymax")] <- eez_bbox[c("xmax", "ymax")] + 2
  land_cropped <- st_crop(land, eez_bbox)
  
  # Create and display map
 map <- tm_shape(land_cropped) +
    tm_polygons(fill = "lightgray", col = "black", lwd = 1) +
    tm_shape(eez_data) +
    tm_polygons(fill = "Suitable Area (km2)", 
                fill.scale = tm_scale_continuous(values = "brewer.yl_or_rd"),
                fill.legend = tm_legend(title = "Suitable Area (km²)"),
                col = "black",
                lwd = 0.5) +
    tm_text("rgn", size = 0.4, col = "black", fontface = "bold", 
            remove.overlap = TRUE, auto.placement = TRUE) +
    tm_title(text = paste("Most Suitable EEZ for", species_name), 
             size = 0.7, fontface = "bold") +
    tm_scalebar(position = c("left", "bottom")) +
    tm_compass(position = c("right", "top")) +
    tm_layout(legend.outside = TRUE, legend.outside.position = "right")
  
 # Clean Environment After running Function
 rm(sst_rcl, depth_rcl, sst_bin, depth_bin, depth_stack, suitable_cells, eez_data, suitable_masked, eez_rast, cell_area, suitable_area_rast, total_area_per_eez, ranked_eez, land, eez_bbox, land_cropped)
 
  print(map)
  return(invisible(map))
}
```

# Most suitable EEZ for Oysters

```{r, warning=FALSE, message=FALSE}
# Finding the most suitable EEZ habitats for Oysters
map_species_suitability("Oysters", sst_min = 11, sst_max = 30, depth_min = -70, depth_max = 0)
```

The most suitable Exclusive Economic Zone for Oysters generally is Central California.

# Most suitable EEZ for Olympia oysters

Data: <https://www.sealifebase.ca/summary/Ostrea-lurida.html>

```{r, warning=FALSE, message=FALSE}
# Finding the most suitable EEZ habitats for Olympia Oysters
map_species_suitability("Olympia Oysters", sst_min = 10, sst_max = 12, depth_min = -71, depth_max = 0)
```

The most suitable Exclusive Economic Zone for Olympia Oysters is Washington.

# Most suitable EEZ for Red Abalone

Data: [https://www.sealifebase.ca/summary/Haliotis-rufescens.html](https://www.sealifebase.ca/summary/Haliotis-rufescens.html%5D)

```{r, warning=FALSE, message=FALSE}
# Finding the most suitable EEZ habitats for Red Abalone
map_species_suitability("Red Abalone", sst_min = 8, sst_max = 18, depth_min = -24, depth_max = 0)
```

The most suitable Exclusive Economic Zone for Red Abalone is Washington.
